---
title: 'Introduction to Bayesian GLMs'
always_allow_html: true
output:
  html_document:
    collapsed: no
    smooth_scroll: no
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '4'
---  

# R course: Introduction to Bayesian GLMs

[Chris Brown](https://experts.griffith.edu.au/7867-chris-brown)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results ='hide', warnings = FALSE, message = FALSE, comment = '', strip.white = FALSE)
```  

Welcome to  "Introduction to Bayesian GLMs"

# Agenda 

    Intro to Bayesian stats theory

    GLM basics and simulating data 
    
    Simple GLM with brms package

    Checking convergence

    Plotting posteriors

    Lunch break ~ 30-45 minutes

    Under the hood - custom stan code

    Prior predictive checks 

    Random effects in stan code 
    
    Random effects with brms 
    
    Importance of thinking about variance priors 

    Non-linear models 
    
    Many things you can do with brms

    Conclusion 

# About you 

Introductions, what do you want to use Bayesian stats for? 
Review quiz answers


# Intro to Bayesian stats theory

Probability theory  

`P(A)`  

`P(A|B)`  

`P(A|B)*P(B) = P(B|A)*P(A)`  

`P(A,B) = P(A|B)*P(B)`

Now pretend A and B are data and models 

Frequentist statistics 

P(D|M) = Likelihood

The model can be one parameter value e.g. a mean

We look for the maximuum likelihood. 

Hence the idea that there is a fixed 'truth' for frequentists


Bayesians interpret it this way:

P(M|D) = P(D|M)*P(M)/P(D)

Explain each of these terms 

Hence the idea that the truth is a random variable. 
[Disease example](https://towardsdatascience.com/intro-to-bayesian-statistics-5056b43d248d)

What is P(D) though?  

It's a normalizing constant. Here's a simple example of how we can solve for that
Assume there are only two models therefore

`P(D) = P(D, M1) + P(D, M2)` 

`P(D) = P(D|M1)P(M1) + P(D|M2)P(M2)`

`P(M|D) = P(D|M)*P(M)/[P(D|M1)P(M1) + P(D|M2)P(M2)]`

If there are three models the denominator has three additions and so on. 

If the model space is continuous (ie a mean parameter) then we use integration to solve for P(D)
Integration for complex multiparameter models can get very nasty...
So we use an algorithm to do that. 
Enter MCMC. We'll explore this later on. 
But for know just know that the `P(D)` term has no practical meaning (unlike the other terms) and is just a constant that ensures the LHS sums (or integrates) to 1. 

Case study of owls 

A = barking owl

B = observed owl


P(owl/obs) = p(obs/there)p( owl is there) / p(obs)
P(owl) = 0.005
P(obs/there) = 0.5
P(obs/not there) = 0.01(mistake for boobook)
P(obs) = sum prob of all possible ways of seeing an owl =
Prob no owl but see it + prob owl and we do see it =
`0.995*0.01+ 0.5*0.005`

So overall probability of barking owl given my observation = 
```{r}
0.5*0.005/((0.995*0.01) +(0.5*0.005))
```

Bayesian thinking can be a way of life.  

# Case-study fish vs SST 

Regression of fish vs SST
What do you know about if this is all I told you. 
What are our priors? 
Let's do it as frequentists first, then as bayesians. 

# Maximum likelihood GLM

```{r}
library(brms)
library(dplyr)
library(ggplot2)
library(patchwork)
library(tidyr)
theme_set(theme_classic())
```


## Generate some data 

Sorry not using simstudy anymore! 

```{r}
pois_glm_sim <- function(params, x, seed = runif(1, 0, 100000)){
    set.seed(seed)
  
  with(params,{
    lambda = exp(alpha + beta*x)
    y = rpois(length(x), lambda)
    data.frame(id = 1:length(y), alpha, beta, x, lambda, y)
  })
    
}

```


```{r}
set.seed(55)
N <- 50
params <- data.frame(alpha = 0,
               beta = 0.15)
sst_range <- c(8, 25)
sst <- runif(N, sst_range[1], sst_range[2])
sstnorm <- scale(sst)
dat <- pois_glm_sim(params, x = sst, seed = 42)

```

```{r}
ggplot(dat) + 
  aes(x = x, y = y) + 
  geom_point()
```

## Fitting a glm

```{r}
m1 <- glm(y ~ x, family = "poisson",
          data = dat)
pred_m1 <- predict(m1,
                        se = TRUE)
coef(m1)
```

```{r}
dat$pred_glm <- pred_m1$fit
dat$pred_glm_se <- pred_m1$se.fit

g1 <- ggplot(dat) + 
  aes(x = x, y = y) + 
  geom_point() + 
  geom_line(aes(y = exp(pred_glm)), color = "red") + 
  geom_ribbon(aes(ymin = exp(pred_glm - pred_glm_se*1.96),
                  ymax = exp(pred_glm + pred_glm_se*1.96)),
              alpha = 0.5, 
              fill = "red")

g1
```  

# Bayesian GLM with brms

Try this first with only 1050 samples

```{r eval=FALSE}
options(mc.cores = parallel::detectCores())

b1 <- brm(y ~ x, family = "poisson",
          data = dat,
          chains = 4,
          iter = 2000,
          thin = 1,
          warmup = 1000,
          cores = 4)
# save(b1, file = "models/b1.rda")

```
```{r echo=FALSE}
load("models/b1.rda")
```

```{r}
summary(b1)
```

## Checking convergence 

```{r}
plot(b1)
```


```{r eval = FALSE}
# shinystan::launch_shinystan(b1)
```

#### The folk theorem of statistical computing

"When you have computational problems, often there’s a problem with your model (Yao, Vehtari, and Gelman, 2020)" (or your model is misspecified with respect to the data at hand)

## Plotting predictions 

```{r}
conditional_effects(b1)
```

## Extracting samples 

```{r}
pred_b1 <- posterior_epred(b1)
dim(pred_b1)

```

```{r}
quantile(pred_b1[,1], c(0.025, 0.5, 0.975))
```

# Comparing to GLM 


```{r}
dat$pred_brm <- colMeans(pred_b1)
dat$pred_brm_lwr <- apply(pred_b1, 2, 
                          quantile, probs = 0.025)
dat$pred_brm_upr <- apply(pred_b1, 2, 
                          quantile, probs = 0.975)

g2 <- ggplot(dat) + 
  aes(x = x, y = y) + 
  geom_point() + 
  geom_line(aes(y = pred_brm), color = "blue") + 
  geom_ribbon(aes(ymin = pred_brm_lwr,
                  ymax = pred_brm_upr),
              alpha = 0.5, 
              fill = "blue")
```

```{r}
g1 + g2
```

# Predictive intervals 

```{r}
pred_b1 <- posterior_predict(b1)
dat$pred_int <- colMeans(pred_b1)
dat$pred_int_lwr <- apply(pred_b1, 2, 
                          quantile, probs = 0.025)
dat$pred_int_upr <- apply(pred_b1, 2, 
                          quantile, probs = 0.975)

g3 <- ggplot(dat) + 
  aes(x = x, y = y) + 
  geom_point() + 
  geom_line(aes(y = pred_int), color = "green") + 
  geom_ribbon(aes(ymin = pred_brm_lwr,
                  ymax = pred_brm_upr),alpha = 0.5, 
              fill = "green") +
  geom_ribbon(aes(ymin = pred_int_lwr,
                  ymax = pred_int_upr),
              alpha = 0.5, 
              fill = "lightblue")
g3

```

# Under the hood - stan code 

## Simulating data from priors 

AKA prior predictive check 

```{r}
library(rstan)
options(mc.cores = parallel::detectCores())

dat_in <- list(N = nrow(dat), y = dat$y,
               x = dat$x)

compiled_model <- stan_model("glm-poisson-sim.stan")

sim_out <- sampling(
  compiled_model, data = dat_in)

prior_sims  <- sim_out %>% 
  as.data.frame %>% 
  select(contains("y_sim"))

summary_tbl <- apply(prior_sims[1:5,], 1, summary)
summary_tbl
```

```{r}
exp(1 + max(dat$x)*1)
```
Wow!

Try normalize x  
```{r}
dat$xscale <- scale(dat$x)
exp(4+ max(dat$xscale)*2)
```
```{r}
dat_in <- list(N = nrow(dat), y = dat$y,
               x = as.numeric(dat$xscale))

sim_out <- sampling(
  compiled_model, data = dat_in)

prior_sims  <- sim_out %>% 
  as.data.frame %>% 
  select(contains("y_sim"))

summary_tbl <- apply(prior_sims[1:5,], 1, summary)
summary_tbl
```

```{r}
sims_dat <- prior_sims[1:15,] %>% 
  t() %>%
  data.frame() %>%
  mutate(x = dat_in$x) %>%
  pivot_longer(-x, names_to = "sim",
               values_to = "y")
  
ggplot(sims_dat) +
  aes(x = x,y = y) + 
  geom_point() + 
  facet_wrap(~sim, scale = "free")
```

## Fitting the stan model 


```{r, message=FALSE,warning=FALSE,eval=FALSE}

dat_in <- list(N = nrow(dat), y = dat$y,
               x = as.numeric(dat$xscale))

fit1 <- stan(file = "glm-poisson.stan",
          data = dat_in,
          iter = 2000,
          warmup = 1000,
          thin = 1,
          chains =  4,
          refresh = 500,
          )
save(fit1, file = "models/fit1.rda")

```
```{r echo=FALSE}
load("models/fit1.rda")
```

```{r}
xout <- rstan::extract(fit1)
names(xout)
quantile(xout$alpha, c(0.025, 0.5, 0.975))
quantile(xout$beta, c(0.025, 0.5, 0.975))
plot(density(xout$alpha))
```

```{r}
fit_summary <- summary(fit1, pars = c("alpha", "beta"),
                       probs = c(0.033, 0.5, 0.67))
fit_summary$summary
```
## Plotting predictions 

We can use the `y_sim` or do it this way: 

Make a list of dataframes: 

```{r}
listparams <- Map(function(alpha,beta) data.frame(alpha,
                              beta), 
       xout$alpha,
       xout$beta)
```

### Predictions using our function 

```{r}
datpred <- lapply(listparams, pois_glm_sim,
                  x = dat$xscale) %>%
  do.call("rbind", .)

datpred <- datpred %>%
  group_by(id) %>%
  summarize(x = mean(x),
            lambda = mean(lambda))

ggplot(datpred) + 
  aes(x = x, y = lambda) + 
  geom_line()
```


# Random effects models 

```{r}
pois_glmm_sim <- function(params, x, 
                          N_per_group, 
                          ngrps,
                          seed = runif(1,0, 10000)){
  
  set.seed(seed)
  
  #total sample size
  N <- N_per_group * ngrps
  
  with(params, {
    #random effects 
    z <- rnorm(ngrps, 0, sigma)
    Z <- rep(z, each = N_per_group)

    #poisson mean
    lambda <- exp(alpha + beta*x + Z)
    y <- rpois(N, lambda)
    data.frame(id = 1:length(y), 
               x, lambda, y,
               #just a trick to convert Z into integer site IDs:
               siteID = as.numeric(factor(Z)))
  })
  
}
```


```{r}
#sst values (1 per group/site)
ngrps <- 12
N_per_group <- 5
x <- runif(ngrps, sst_range[1], sst_range[2])
x <- scale(x)
x <- rep(x, each = N_per_group) 
    
```


```{r}
params <- list(alpha = 2.5,
               beta = 0.75,
               sigma = 0.3)

set.seed(77)
dat2 <- pois_glmm_sim(params, x,
                          N_per_group, ngrps, 
                          77)


```

try varying sigma!

```{r}
ggplot(dat2) + 
  aes(x = x, y = y, color = factor(siteID)) + 
  geom_point()
```

## Random effects with stan  

```{r, message=FALSE,warning=FALSE,eval=FALSE}

dat_in <- list(N = nrow(dat2), y = dat2$y,
               x = dat2$x,
               igroup = dat2$siteID,
               K = length(unique(dat2$siteID)))

fitRE <- stan(file = "glm-poisson-random-effects.stan",
          data = dat_in,
          iter = 2000,
          warmup = 1000,
          thin = 1,
          chains =  4,
          refresh = 500)
save(fitRE, file = "models/fitRE.rda")
```

```{r echo=FALSE}
load("models/fitRE.rda")
```

```{r}
fit_summary <- summary(fitRE, pars = c("alpha", "beta", "sigma"),
                       probs = c(0.033, 0.5, 0.67))
fit_summary$summary
```
And compare to what we put in: 

```{r}
params
```
## Random effects with brms  

```{r, message=FALSE,warning=FALSE,eval=FALSE}
b1ME <- brm(y ~ x + (1|siteID), family = "poisson",
          data = dat2,
          chains = 4,
          iter = 2000,
          thin = 1,
          warmup = 1000,
          cores = 4)
save(b1ME, file = "models/b1ME.rda")

```

```{r echo=FALSE}
load("models/b1ME.rda")
```

```{r}
summary(b1ME)
```


```{r}
plot(b1ME)
```

```{r}
conditional_effects(b1ME) 
```

### Look at random effects estimates

```{r}
b1z <- ranef(b1ME, groups = "siteID")$siteID %>%
  data.frame() %>%
  mutate(i = 1:n())

ggplot(b1z) + 
  aes(x = i) + 
  geom_point(aes(y = Estimate.Intercept)) + 
  geom_linerange(aes(ymin = Q2.5.Intercept,
                     ymax = Q97.5.Intercept))
```

[Check normality with qqnorm plots](https://www.seascapemodels.org/rstats/2017/10/06/qqplot-non-normal-glm.html)

```{r}
qqnorm(b1z$Estimate.Intercept)
qqline(b1z$Estimate.Intercept, lty = 2)
```

# Importance of variance priors 

```{r}
hist(rgamma(10000, 10, 0.001))
```


```{r, message=FALSE,warning=FALSE}
b1_priors <- get_prior(y ~ x + (1|siteID), family = "poisson",
          data = dat2)
b1_priors
prior_sigma <- set_prior("gamma(10, 0.001)", 
                         class = "sd", coef = "Intercept",
                         group = "siteID")
make_stancode(y ~ x + (1|siteID), family = "poisson",
          data = dat2,
          prior = prior_sigma,
          save_model = "dumb-prior.stan")
```


```{r, message=FALSE,warning=FALSE,eval=FALSE}
b1ME_gamma001 <- brm(y ~ x + (1|siteID), family = "poisson",
          data = dat2,
          chains = 4,
          iter = 2000,
          thin = 1,
          warmup = 1000,
          cores = 4,
          prior = prior_sigma)
save(b1ME_gamma001, file = "models/b1ME_gamma001.rda")

```

```{r echo=FALSE}
load("models/b1ME_gamma001.rda")
```

```{r}
summary(b1ME_gamma001)
```


```{r}
plot(b1ME_gamma001)
```

Notice the chains never converge! 

```{r eval = FALSE}
# shinystan::launch_shinystan(b1)
```


# Non-linear models in BRMS 

```{r}
pois_gamm_sim <- function(params, x, 
                          N_per_group, 
                          ngrps,
                          seed = runif(1,0, 10000)){
  
  set.seed(seed)
  
  #total sample size
  N <- N_per_group * ngrps
  
  with(params, {
    #random effects 
    xs <- splines::ns(x,3)
    z <- rnorm(ngrps, 0, sigma)
    Z <- rep(z, each = N_per_group)

    #poisson mean
    lambda <- exp(alpha + beta1*xs[,1] + beta2*xs[,2] + beta3*xs[,3] + Z)
    y <- rpois(N, lambda)
    data.frame(id = 1:length(y), 
               x, lambda, y,
               #just a trick to convert Z into integer site IDs:
               siteID = as.numeric(factor(Z)))
  })
  
}
```

```{r}
params <- list(alpha = 1,
               beta1 = -1,
               beta2 = 3,
               beta3 = -1,
               sigma = 0.2)

set.seed(77)
dat3 <- pois_gamm_sim(params, x,
                          N_per_group, ngrps, 
                          77)

ggplot(dat3) + 
  aes(x = x, y = y, color = factor(siteID)) + 
  geom_point()
```

```{r}
b1gam1 <- brm(y ~ s(x) + (1|siteID), family = "poisson",
          data = dat3,
          chains = 4,
          iter = 2000,
          thin = 1,
          warmup = 1000,
          cores = 4)
save(b1gam1, file = "models/b1gam1.rda")

```

```{r echo=FALSE}
load("models/b1gam1.rda")
```

```{r}
summary(b1gam1)
```


```{r}
plot(b1gam1)
```

```{r}
conditional_effects(b1gam1)
```

# Other ideas for things you can do with brms 

- Hurdle models (for zero inflated data )  

- Simple multivariate models  

- Model missing values of fixed effects   

- Measurement error models/meta-analyses  

- Splines  

- Simple state space models (e.g. AR)  

- Ordinal/categorical models  

- Models where variance/dispersian is regressed against parameters 

- Multilevel models (e.g. RE*fixed effect interactions)

And much more! 

# Priors for fixed effects 

```{r eval=FALSE}

tighter_prior <- 
  set_prior("normal(-5,0.1)", class = "b", coef = "sst")
b2 <- brm(y ~ x, family = "poisson",
          data = dat,
          chains = 4,
          iter = 2000,
          thin = 1,
          warmup = 1000,
          cores = 4,
          prior = tighter_prior)
summary(b2)
```

### Comparing to GLM 


```{r eval=FALSE}
pred_b2 <- posterior_epred(b2)
dat$pred_brm2 <- colMeans(pred_b2)
dat$pred_brm_lwr2 <- apply(pred_b2, 2, 
                          quantile, probs = 0.025)
dat$pred_brm_upr2 <- apply(pred_b2, 2, 
                          quantile, probs = 0.975)

ggplot(dat) + 
  aes(x = sst, y = fish) + 
  geom_point() +
  geom_line(aes(y = pred_brm), color = "blue") + 
  geom_ribbon(aes(ymin = pred_brm_lwr,
                  ymax = pred_brm_upr),
              alpha = 0.5, 
              fill = "blue") + 
  geom_line(aes(y = pred_brm2), color = "green") + 
  geom_ribbon(aes(ymin = pred_brm_lwr2,
                  ymax = pred_brm_upr2),
              alpha = 0.5, 
              fill = "green")
  
  
```

Use predict(b1) if you want the predictive intervals 


# Practical things

[Stan userguide online](https://mc-stan.org/users/documentation/)

[Stan and brms forums for Q&A](https://discourse.mc-stan.org/)

My favourite introductory book: [Statistical Rethinking](https://www.amazon.com.au/s?k=statistical+rethinking&adgrpid=88123238558&gclid=CjwKCAiAs92MBhAXEiwAXTi256pXVqV0J16VwZFOn_u_ig2jlQPfGBCeJ3qkSDDZiA4UAD3Rx4sokBoCAGwQAvD_BwE&hvadid=518058673165&hvdev=c&hvlocphy=9069204&hvnetw=g&hvqmt=e&hvrand=10575411477101464194&hvtargid=kwd-300572196203&hydadcr=1924_218846&tag=googhydr0au-22&ref=pd_sl_7w90ddlgy2_e)

[Bayesian workflow](https://arxiv.org/abs/2011.01808)

If you need to speed up analyses, try the INLA package (but it works differently, not using MCMC)


